A Linear Solution of the Linearity Challenge

### 2025-12-25 paper/calculus.tex
revision

In this section we give a cursory presentation of \LCP.
%
As we have anticipated in \Cref{sec:introduction}, \LCP is very closely related
to \CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear
channels instead of sessions. A session consisting of several messages exchanged
in sequence can be encoded in \LCP using linear channels by pairing each message
with a fresh continuation channel on which the rest of the session takes place.
%
This way of encoding sessions has been first described by \citet[extended
version]{Kobayashi02b} and later studied by \citet{DardhaGiachinoSangiorgi17}.

@@

The interpretation of propositions as session types is the usual
one~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
protocols that terminate with the input/output of a signal; the additive
connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
continuations $A$ and $B$ with are selected/offered; the multiplicative
connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
continuations $A$ and $B$. These forms are typically used to describe the
input/output of a channel (to be used according to $A$) and then continuing as
$B$. The additive constants $\Zero$ and $\Top$ do not describe useful protocols,
but they can play the role of smallest/largest type in type systems that support
a notion of subtyping~\citep{HornePadovani24}. 
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
type $X$ followed by the protocol $A$ and are useful to describe parametric
protocol polymorphism.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe servers and clients accepting and requesting connections to
be used according to $A$.

The notions of \emph{free type variables}, \emph{duality} and of \emph{type
substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
the set of type variables occurring free in $A$, we write $\dual{A}$ for the
dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the free
occurrences of $X$ in $A$ with $B$.

@@

The syntax of \Calculus makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
every message sent on one of the channels is forwarded to the other. As
discussed in the literature~\cite{LindleyMorris15}, this form is also useful for
modeling delegation, that is the exchange of an existing session endpoint in
another session.
%
The $\Wait\x.P$ and $\Close\x$ processes respectively model the input and output
of a termination signal on the channel $x$. The latter process has no
continuation, while the former process continues as $P$ once the signal has been
received.
%
The process $\Fail\x$ can be used to denote a failing protocol concerning the
channel $x$.
%
The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
describe the input/output of a label $\InTag_i$ along with a fresh continuation
$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
received label $\InTag_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical $\InTag_1$ and
$\InTag_2$. This generalisation has no substantial impact on the formalisation,
so we stick to binary choices in line with the usual presentations of linear
logic.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
forks into $P$ and $Q$, each using $y$ and $z$ respectively. The latter process
continues as $R$. These forms are often used in conjunction with links to model
\emph{delegation}.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

Next we have process forms dealing with shared (non-linear) channels. The
$\Server\x\y.P$ and $\Client\x\y.P$ processes respectively denote \emph{servers}
and \emph{clients} acting on the shared channel $x$. Each request (from a
client) spawns a copy of the server's body using the continuation channel $y$.
The process $\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client
that \emph{does not} use the server waiting on $x$. The process
$\Contract\x\y\z.P$ denotes an explicit \emph{contraction}, that is a client
that connects to the server waiting on $x$ multiple times on the channels $y$
and $z$.

\emph{Cuts} of the form $\Cut[A]\x{P}{Q}$ represent the parallel composition of
the processes $P$ and $Q$ connected by a channel $x$, which has type $A$ in $P$
and type $\dual{A}$ in $Q$. Henceforth we write $\Cut\x{P}{Q}$ omitting the type
annotation $A$ when it is irrelevant or clear from the context.

The notions of free and bound channels are fairly standard, bearing in mind that
output prefixes bind continuation channels in (some) continuation processes. In
particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in $P$.
%
We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
processes up to renaming of bound channels.

@@

    As an example, the composition $\Cut[\Bool]\x{\TrueP(x)}{\NotP(x,y)}$
    produces false on $y$.
    %
    Note the use of explicit continuations in these processes and the fact that
    each channel is used \emph{exactly once}. The same processes in \CP would be
    written as
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{}.\Close\x
        \quad
        \FalseP(x) \eqdef \Select\x\RightTag{}.\Close\x
        \quad
        \NotP(x,y) \eqdef
        \BigCase\x{}{
            \Wait\x.\FalseP(y)
        }{
            \Wait\x.\TrueP(y)
        }
    \]
    where each channel is used multiple times to indicate the various
    input/output actions pertaining to the same session.
    %
    In general, the \CP version of an \LCP process can be obtained by reusing
    the same channel $x$ in place of the continuation $z$ in \Cref{tab:syntax}.
    %
    \eoe
\end{example}

@@

\begin{table}[t]
    \caption{Operational semantics of \Calculus. Some side conditions for the
    \rulename{s-*} rules are omitted (see text).}
    \label{tab:semantics}
    \begin{tabular}{@{}c@{}}
        \begin{math}
            \displaystyle
            \begin{array}{rrcll}
            \SLink &
            \Link\x\y & \pcong & \Link\y\x
            \\
            \SComm &
            \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P}
            \\
            \SFail &
            \Cut\x{\Fail\y}{P} & \pcong & \Fail\y % & x \ne y
            \\
            \SWait &
            \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} % & x \ne y
            \\
            \SCase &
            \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} % & x \ne y,z
            \\
            \SSelect &
            \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SJoin &
            \Cut\x{\Join\y\u\z.P}{Q} & \pcong & \Join\y\u\z.\Cut\x{P}{Q} % & x \ne y,u,v
            \\
            \SForkL &
            \Cut\x{\Fork\y\u\z{P}{Q}}{R} & \pcong & \Fork\y\u\z{\Cut\x{P}{R}}{Q} & x \in \fn{P}
            \\
            \SForkR &
            \Cut\x{\Fork\y\u\z{P}{Q}}{R} & \pcong & \Fork\y\u\z{P}{\Cut\x{Q}{R}} & x \in \fn{Q} %\setminus\set{y,u,v}
            \\
            \SForall &
            \Cut\x{\Forall\y\X\z.P}{Q} & \pcong & \Forall\y\X\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SExists &
            \Cut\x{\Exists\y{A}\z.P}{Q} & \pcong & \Exists\y{A}\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SServer &
            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} % & x \ne y,u
            \\
            \SClient &
            \Cut\x{\Client\y\z.P}{Q} & \pcong & \Client\y\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SWeaken &
            \Cut\x{\Weaken\y.P}{Q} & \pcong & \Weaken\y.\Cut\x{P}{Q} % & x \ne y
            \\
            \SContract &
            \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} % & x \ne y,u,v
            \\
            \\
            \RLink &
            \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
            \RClose &
            \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
            \RSelect &
            \Cut\x{\Select\x{\InTag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
            \RFork &
            \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
            \RExists &
            \Cut\x{\Exists\x{A}\z.P}{\Forall\x\X\z.Q} & \red &
            \Cut\z{P}{Q\subst{A}\X} \\
            \RConnect &
            \Cut\x{\Server\x\y.P}{\Client\x\y.Q} & \red & \Cut\y{P}{Q} \\
            \RWeaken &
            \Cut\x{\Server\x\y.P}{\Weaken\x.Q} & \red & \Weaken{\seqof\z}.Q
            & \fn{P} = \set{y,\seqof\z} \\
            \RContract &
            \Cut\x{\Server\x\y.P}{\Contract\x{x'}{x''}.Q} & \red &
            \multicolumn{2}{l}{
                \Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.\Cut{x'}{
                    \Server{x'}\y.P'
                }{
                    \Cut{x''}{
                        \Server{x''}\y.P''
                    }{
                        Q
                    }
                }
                ~^*
            }
            \\
            \RCut &
            \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
            \RCong &
            P & \red & Q & P \pcong R \red Q \\
            \\
            \multicolumn{5}{c}{
                ~^*
                \fn{P} = \set{y,\seqof\z},
                P' = P\subst{\seqof\z'}{\seqof\z},
                P'' = P\subst{\seqof\z''}{\seqof\z}
            }
            \end{array}
        \end{math}
    \end{tabular}
\end{table}

@@

Structural pre-congruence is the least pre-congruence defined by the
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are commutative.
The remaining rules, when read from left to right, push a cut on $x$ underneath
the topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These
rules are key to float input/output actions to the top-level of a process, so
that they can interact with corresponding complementary actions in the context
surrounding the process.
%
All these rules have implicit side-conditions aimed at preserving the meaning of
channels when binders are moved around: terms entering or exiting the scope of a
binder for $x$ must not have free occurrences of $x$. This holds also for the
type variable $X$ in the rule \SForall.
%
We content ourselves with such informal description of these side conditions
given that we are going to formalise \Calculus later on.
%
Note also that there are two versions of \SForkL and \SForkR depending on which
of the two continuations $P$ and $Q$ contains free occurrences of the restricted
channel $x$.

There is no pre-congruence rule asserting that parallel composition is
associative. While this is a sound and expected property of parallel
composition, it turns out to be inessential for proving the properties of
\Calculus that we consider. We leave it out in light of the forthcoming
formalisation of the calculus, in which every simplification is welcome,
although standard presentations of classical linear logic admit such
law~\cite{Wadler14,Doumane17}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists, \RConnect,
\RWeaken and \RContract are in one-to-one correspondence with the well-known
reductions of session calculi and the principal cut reductions of linear logic.
\RLink eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes
a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
respectively model the communication of a tag $\InTag_i$ and of a fresh channel
$y$. \RExists models the instantiation of a polymorphic variable $X$ as the
communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
replacing every free occurrence of the type variable $X$ with $A$. Each of these
communications consumes the channel $x$ and possibly creates and transfers a
\emph{fresh continuation} $z$ on which the rest of the conversation takes place,
in line with the encoding of sessions into the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
The rules \RConnect, \RWeaken and \RContract describe interactions on shared
channels and respectively represent the connection between a server and a
client, the disposal of an unused server, and the duplication of a server. In
\RWeaken and \RContract we use some convenient (albeit slightly ambiguous)
notation for denoting sequences of (pairwise distinct) channels and prefixes. In
particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

@@

    We establish that $\EchoP(x)$ is well typed with the following deerivation
    \[
        \begin{prooftree}
            \[
                \[
                    \[
                        \[
                            \justifies
                            \wtp{
                                \Link\u\v
                            }{
                                u : X,
                                v : \dual\X
                            }
                            \using\LinkRule
                        \]
                        \[
                            \justifies
                            \wtp{
                                \Close{y'''}
                            }{
                                y''' : \One
                            }
                            \using\CloseRule
                        \]
                        \justifies
                        \wtp{
                            \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                        }{
                            u : X,
                            y'' : \dual\X \Ten \One
                        }
                        \using\ForkRule
                    \]
                    \justifies
                    \wtp{
                        \Join{y'}\u{y''}.
                        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                    }{
                        y' : X \Par (\dual\X \Ten \One)
                    }
                    \using\JoinRule
                \]
                \justifies
                \wtp{
                    \Forall\y\X{y'}.
                    \Join{y'}\u{y''}.
                    \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                }{
                    y : \forall\X.X \Par (\dual\X \Ten \One)
                }
                \using\ForallRule
            \]
            \justifies
            \wtp{
                \EchoP(x)
            }{
                x : \OfCourse(\forall\X.X \Par (\dual\X \Ten \One))
            }
            \using\ServerRule
        \end{prooftree}
    \]
    %
    where the side condition of the rule \ForallRule is trivially satifised
    since the typing context does not contain bindings other than the one for
    $y$.
    %
    \eoe
\end{example}

@@

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalising two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) peer endpoints are always used in complementary ways. This
latter property is called \emph{well formedness} in the challenge. In this work
we also consider \emph{deadlock freedom}, which is more general than well
formedness and holds for \Calculus since its type system is based on linear
logic. We now formulate these properties using the notation developed so far.

@@

In order to formulate deadlock freedom, we first need to introduce some
terminology for referring to processes that are unable to make any progress. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
process is unable to reduce (because a $\Close\x$ process is meant to interact
with a process of the form $\Wait\x.P$) and, more generally, it is unable to
interact regardless of the context in which it occurs because the $\Close\x$
sub-processes it contains are blocked on the channel $x$ that is restricted by
the cut.
%
In general, the property of being a deadlock is not simply the inability to
reduce: there are irreducible processes that are not a deadlock because they
would be able to reduce if put into a suitable context. For example, $\Wait\x.P$
does not reduce, and yet it is not a deadlock because it would be able to make
progress when composed in parallel with $\Close\x$. Even a process like
$\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a deadlock,
because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using \SWait and
possibly \SComm. Let us make all this more precise.

@@

We now shift the attention to \emph{well formedness}. In the linearity
challenge~\cite{CarboneEtAl24}, this property ensures that, whenever two
processes composed in parallel begin with actions concerning the very same
session, then such actions are \emph{complementary}, in the sense that they
describe opposite forms of interaction. To define well formedness in our
setting, we introduce \emph{reduction contexts} as partial processes with a
single unguarded hole $\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut[A]\x\ReductionContext{P} \mid \Cut[A]\x{P}\ReductionContext
\]

>>>>

In this section we give a cursory presentation of \LCP.
%
As we have anticipated in \Cref{sec:introduction}, \LCP is closely related to
\CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear channels
instead of sessions. A session consisting of several exchanged messages is
encoded using linear channels by pairing each message with a fresh continuation
channel on which the rest of the session takes place.
%
This way of encoding sessions has been first described by \citet[extended
version]{Kobayashi02b} and later studied by \citet{DardhaGiachinoSangiorgi17}.

@@

The interpretation of linear logic propositions as protocols is fairly
usual~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
protocols that terminate with the input/output of a signal; the additive
connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
continuations of type $A$ and $B$ which are selected/offered; the multiplicative
connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
continuations of type $A$ and $B$. The additive constants $\Zero$ and $\Top$ do
not describe useful protocols, but they can play the role of smallest/largest
type in type systems that support a notion of subtyping~\citep{HornePadovani24}. 
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
type $X$ with a continuation $A$ and are useful to describe parametric protocol
polymorphism.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe shared channels on which servers and clients accept and
request connections of type $A$.

The notions of \emph{free type variables}, of \emph{duality} and of \emph{type
substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
the set of type variables occurring free in $A$, we write $\dual{A}$ for the
dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the free
occurrences of $X$ in $A$ with $B$.

@@

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
every message sent on one of the channels is forwarded to the other. As
discussed in the literature~\cite{LindleyMorris15}, this form is useful for
modeling higher-order sessions, that is the exchange of an existing channel on
another channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of a termination signal on the channel $x$. The latter process has no
continuation, while the former process continues as $P$ once the signal has been
received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
describe the input/output of a label $\InTag_i$ along with a fresh continuation
$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
received label $\InTag_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of labels instead of the canonical $\InTag_1$ and
$\InTag_2$. This generalisation has no substantial impact on the formalisation,
so we stick to binary choices in line with the usual presentations of linear
logic.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
has a single continuation $R$ that can use both $y$ and $z$. The latter process
forks into $P$ and $Q$, each using $y$ and $z$ respectively. These forms are
often used in conjunction with links to model higher-order sessions.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

Next we have process forms dealing with shared (non-linear) channels. The
$\Server\x\y.P$ and $\Client\x\y.P$ processes respectively denote \emph{servers}
and \emph{clients} acting on the shared channel $x$. Each request (from a
client) spawns a copy of the server's body using the continuation channel $y$.
The process $\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client
that \emph{does not} use $x$. The process $\Contract\x\y\z.P$ denotes an
explicit \emph{contraction} whereby a client uses $x$ multiple times (once with
name $y$ and once with name $z$).

Finally, cuts of the form $\Cut[A]\x{P}{Q}$ represent the parallel composition
of the processes $P$ and $Q$ connected by a channel $x$, which has type $A$ in
$P$ and type $\dual{A}$ in $Q$. Henceforth we write $\Cut\x{P}{Q}$ omitting the
type annotation $A$ when it is irrelevant or clear from the context.

The notions of free and bound channels are fairly standard, bearing in mind that
output prefixes bind continuation channels in (some) continuation processes. In
particular, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while
$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
$P$.
%
We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
processes up to renaming of bound channels.

@@

    As an example, the composition $\Cut[\Bool]\x{\TrueP(x)}{\NotP(x,y)}$
    produces false on $y$.
    %
    Note the use of explicit continuations in these processes and the fact that
    each channel is used exactly once. The same processes in \CP would be
    written as
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{}.\Close\x
        \quad
        \FalseP(x) \eqdef \Select\x\RightTag{}.\Close\x
        \quad
        \NotP(x,y) \eqdef
        \BigCase\x{}{
            \Wait\x.\FalseP(y)
        }{
            \Wait\x.\TrueP(y)
        }
    \]
    where each channel is used multiple times to indicate the various
    input/output actions pertaining to the same session.
    %
    In general, the \CP version of an \LCP process can be obtained by reusing
    the same channel $x$ in place of the continuation $z$ in \Cref{tab:syntax}.
    %
    \eoe
\end{example}

@@

\begin{table}[t]
    \caption{Operational semantics of \Calculus. Many side conditions for the
    \rulename{s-*} rules are omitted (see text).}
    \label{tab:semantics}
    \begin{tabular}{@{}c@{}}
        \begin{math}
            \displaystyle
            \begin{array}{rrcll}
            \SLink &
            \Link\x\y & \pcong & \Link\y\x
            \\
            \SComm &
            \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P}
            \\
            \SFail &
            \Cut\x{\Fail\y}{P} & \pcong & \Fail\y % & x \ne y
            \\
            \SWait &
            \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} % & x \ne y
            \\
            \SCase &
            \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} % & x \ne y,z
            \\
            \SSelect &
            \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SJoin &
            \Cut\x{\Join\y\u\z.P}{Q} & \pcong & \Join\y\u\z.\Cut\x{P}{Q} % & x \ne y,u,v
            \\
            \SForkL &
            \Cut\x{\Fork\y\u\z{P}{Q}}{R} & \pcong & \Fork\y\u\z{\Cut\x{P}{R}}{Q} & x \in \fn{P}
            \\
            \SForkR &
            \Cut\x{\Fork\y\u\z{P}{Q}}{R} & \pcong & \Fork\y\u\z{P}{\Cut\x{Q}{R}} & x \in \fn{Q} %\setminus\set{y,u,v}
            \\
            \SForall &
            \Cut\x{\Forall\y\X\z.P}{Q} & \pcong & \Forall\y\X\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SExists &
            \Cut\x{\Exists\y{A}\z.P}{Q} & \pcong & \Exists\y{A}\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SServer &
            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} % & x \ne y,u
            \\
            \SClient &
            \Cut\x{\Client\y\z.P}{Q} & \pcong & \Client\y\z.\Cut\x{P}{Q} % & x \ne y,z
            \\
            \SWeaken &
            \Cut\x{\Weaken\y.P}{Q} & \pcong & \Weaken\y.\Cut\x{P}{Q} % & x \ne y
            \\
            \SContract &
            \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} % & x \ne y,u,v
            \\
            \\
            \RLink &
            \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
            \RClose &
            \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
            \RSelect &
            \Cut\x{\Select\x{\InTag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
            \RFork &
            \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
            \RExists &
            \Cut\x{\Exists\x{A}\z.P}{\Forall\x\X\z.Q} & \red &
            \Cut\z{P}{Q\subst{A}\X} \\
            \RConnect &
            \Cut\x{\Server\x\y.P}{\Client\x\y.Q} & \red & \Cut\y{P}{Q} \\
            \RWeaken &
            \Cut\x{\Server\x\y.P}{\Weaken\x.Q} & \red & \Weaken{\seqof\z}.Q
            & \fn{P} = \set{y,\seqof\z} \\
            \RContract &
            \Cut\x{\Server\x\y.P}{\Contract\x{x'}{x''}.Q} & \red &
            \multicolumn{2}{l}{
                \Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.\Cut{x'}{
                    \Server{x'}\y.P'
                }{
                    \Cut{x''}{
                        \Server{x''}\y.P''
                    }{
                        Q
                    }
                }
                ~^*
            }
            \\
            \RCut &
            \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
            \RCong &
            P & \red & Q & P \pcong R \red Q \\
            \\
            \multicolumn{5}{c}{
                ~^*
                \fn{P} = \set{y,\seqof\z},
                P' = P\subst{\seqof\z'}{\seqof\z},
                P'' = P\subst{\seqof\z''}{\seqof\z}
            }
            \end{array}
        \end{math}
    \end{tabular}
\end{table}

@@

Structural pre-congruence is the least pre-congruence defined by the
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are commutative.
The remaining rules, when read from left to right, push a cut on $x$ underneath
the topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These
rules are key to float input/output actions to the top-level of a process, so
that they can interact with corresponding complementary actions in the context
surrounding the process.
%
All these rules have implicit side-conditions (not shown in
\Cref{tab:semantics}) aimed at preserving the meaning of channels when binders
are moved around: terms entering or exiting the scope of a binder for $x$ must
not have free occurrences of $x$. This holds also for the type variable $X$ in
the rule \SForall.
%
We content ourselves with such informal description of these side conditions
given that we are going to formalise \Calculus later on.
%
Note also that there are two versions of \SForkL and \SForkR depending on which
of the two continuations (either $P$ or $Q$) contains free occurrences of the
restricted channel $x$.

There is no pre-congruence rule asserting that parallel composition is
associative. While this is a sound property of parallel composition, it turns
out to be unnecessary for proving the properties of \LCP that we consider. We
leave it out in light of the forthcoming formalisation of the calculus, in which
every simplification is welcome, although standard presentations of classical
linear logic admit such law~\cite{Wadler14}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists, \RConnect,
\RWeaken and \RContract are in one-to-one correspondence with the well-known
reductions of session calculi and the principal cut reductions of linear logic.
\RLink eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes
a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
respectively model the communication of a tag $\InTag_i$ and of a fresh channel
$y$. \RExists models the instantiation of a polymorphic variable $X$ as the
communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
replacing every free occurrence of the type variable $X$ with $A$. Each of these
communications consumes the channel $x$ and possibly creates and transfers a
\emph{fresh continuation} $z$ on which the rest of the conversation takes place,
in line with the encoding of sessions into the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
The rules \RConnect, \RWeaken and \RContract describe interactions on shared
channels and respectively represent the connection between a server and a
client, the disposal of an unused server, and the duplication of a server. In
\RWeaken and \RContract we use some convenient (albeit slightly informal)
notation for denoting sequences of (pairwise distinct) channels and prefixes. In
particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

@@

    We establish that $\EchoP(x)$ is well typed with the following derivation
    \[
        \begin{prooftree}
            \[
                \[
                    \[
                        \[
                            \justifies
                            \wtp{
                                \Link\u\v
                            }{
                                u : X,
                                v : \dual\X
                            }
                            \using\LinkRule
                        \]
                        \[
                            \justifies
                            \wtp{
                                \Close{y'''}
                            }{
                                y''' : \One
                            }
                            \using\CloseRule
                        \]
                        \justifies
                        \wtp{
                            \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                        }{
                            u : X,
                            y'' : \dual\X \Ten \One
                        }
                        \using\ForkRule
                    \]
                    \justifies
                    \wtp{
                        \Join{y'}\u{y''}.
                        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                    }{
                        y' : X \Par (\dual\X \Ten \One)
                    }
                    \using\JoinRule
                \]
                \justifies
                \wtp{
                    \Forall\y\X{y'}.
                    \Join{y'}\u{y''}.
                    \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                }{
                    y : \forall\X.X \Par (\dual\X \Ten \One)
                }
                \using\ForallRule
            \]
            \justifies
            \wtp{
                \EchoP(x)
            }{
                x : \OfCourse(\forall\X.X \Par (\dual\X \Ten \One))
            }
            \using\ServerRule
        \end{prooftree}
    \]
    %
    where the side condition of the rule \ForallRule is trivially satifised
    since the typing context does not contain bindings other than the one for
    $y$.
    %
    \eoe
\end{example}

@@

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalising two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) the peer endpoints of the same session are always used in
complementary ways. This latter property is called \emph{well formedness} in the
challenge. In this work we also consider \emph{deadlock freedom}, which is more
general than well formedness and holds for \Calculus since its type system is
based on linear logic. We now formulate these properties using the notation
developed so far.

@@

In order to formulate deadlock freedom, we first need to introduce some
terminology for referring to processes that are unable to make any progress. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
process is unable to reduce (because a process $\Close\x$ is meant to interact
with a process of the form $\Wait\x.P$) and, more generally, it is unable to
interact regardless of the context in which it occurs because the sub-processes
it contains are blocked on the channel $x$ that is restricted by the cut.
%
In general, the property of being a deadlock is not simply the inability to
reduce: there are irreducible processes that are not a deadlock because they
would be able to reduce if put into a suitable context. For example, $\Wait\x.P$
does not reduce, and yet it is not a deadlock because it would be able to make
progress when composed in parallel with $\Close\x$. Even a process like
$\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a deadlock,
because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using \SWait and
possibly \SComm. Let us make all this more precise.

@@

We now shift the attention to \emph{well formedness}. In the linearity
challenge~\cite{CarboneEtAl24} this property ensures that, whenever two
processes composed in parallel begin with actions concerning the very same
session, then such actions complement each other, in the sense that they
describe opposite forms of interaction. To define well formedness in our
setting, we introduce \emph{reduction contexts} as partial processes with a
single unguarded hole $\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut[A]\x\ReductionContext{P} \mid \Cut[A]\x{P}\ReductionContext
\]
